---
title: "Calibrating Gene-Set Scores in Your Data"
author: "sparseGenSetCal"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Calibrating Gene-Set Scores in Your Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE, comment = "#>",
    fig.width = 6.5, fig.height = 3.4, dpi = 120,
    fig.align = "center"
)
suppressPackageStartupMessages(library(sparseGenSetCal))
```

```{r, include = FALSE}
# The vignette has to build in seconds, so the permutation counts are turned
# down from their defaults and the matrix is small.  Every number the vignette
# reports is a property of the data, not of these settings, which is the point
# the text makes in the next section; the defaults are stated where they differ.
set.seed(2024)
```

## The question this package answers

You have a gene set, a single-cell matrix, and a hypothesis that the set's
score tracks something -- a cluster, a treatment, a clinical axis. You compute
AUCell scores, you test the correlation, you get a p-value. Before you believe
it, there is one question worth answering: **how much of that score was
measured, and how much of it was the ranking filling in for genes the cell
never expressed?**

That is not a rhetorical question, and its answer is computable in closed form.
Every rank-based score asks where a gene sits among *all* genes in a cell. A
cell that expresses fewer genes than the score's rank ceiling has no count to
place in the remaining positions, so those positions go to genes the cell never
expressed, ordered by whatever rule the implementation uses to break ties. One
such gene contributes, in expectation,

$$\tau(D) = \frac{(m-D)(m-D+1)}{2\,m\,(G-D)},$$

with $m$ the rank ceiling, $G$ the panel size and $D$ the number of genes the
cell detected. It is a function of the cell's depth and not of the gene set --
and depth tracks almost every biological axis one cares about. A score carried
by that term will correlate with the axis for reasons that have nothing to do
with the biology.

This vignette walks through the four steps that replace the null the data
cannot support with one it can, on data small enough to run in a few seconds.
Everything here is an export of the package, so you can paste it into your own
script and point it at your matrix.

## A matrix to work with

`make_synthetic_matrix()` produces Poisson counts with a realistic
gene-activity gradient -- three orders of magnitude between the most and least
active genes -- and a depth distribution you control through `density`. Using
your own data instead means passing the matrix and its gene names; nothing else
changes.

```{r}
X <- make_synthetic_matrix(n_cells = 600, n_genes = 6000, density = 0.04,
                           seed = 11)
dim(X$matrix)
```

The rank ceiling is a fraction of the panel, as in AUCell's default. The cache
holds each gene's rank in each cell whenever that rank is at or below the
ceiling, and zero otherwise; ranks above the ceiling contribute to none of the
scores implemented here, so keeping them would only cost memory.

```{r}
rank_frac <- 0.05
ceiling <- as.integer(ceiling(rank_frac * ncol(X$matrix)))
cache <- rank_cache(X$matrix, X$genes, ceiling = ceiling, seed = 2024)
cache
```

One number decides whether any of this matters, and it can be read off the
depth distribution before a gene set is chosen: the ceiling divided by the
median number of genes a cell detects. Below one, every rank inside the ceiling
holds a measured count and the tie-break never enters the score. Above one, the
slots the cell cannot fill are given to genes it never expressed.

```{r}
median_depth <- stats::median(cache$depth)
data.frame(quantity = c("max_rank", "median genes detected", "ratio"),
           value = c(ceiling, median_depth, round(ceiling / median_depth, 2)))
```

Here the ratio is above one, which is the regime the whole framework is about.
It is also the common one: a shallow cell in a large panel cannot fill a
top-5% block with genes it detected.

## Step 1: how much of the score was never measured

`sparsity_report()` needs the cache and the set, and nothing else. It reports
the share of cells scoring exactly zero, the detection profile of the set, and
-- given the depth vector, which the cache carries -- the share of the set's
chance of entering the score that the tie-break supplies.

The thresholds the verdict uses are visible in the table below, which walks a
set through detection rates from 0.2% to 80%. Real gene sets are not more
exotic than this: a set of transcription factors and a set of ribosomal
proteins are the two ends of it.

```{r}
set_at_detection <- function(target, n = 10) {
    cache$genes[order(abs(cache$detection - target))][seq_len(n)]
}
gradient <- function(target) {
    r <- sparsity_report(cache, set_at_detection(target), rank_frac = rank_frac)
    v <- verdict(r)
    data.frame(detection = target,
               median_detection = r$median_detection,
               zero_rate = r$observed_zero_rate,
               share = r$tie_break_share,
               verdict = v$verdict,
               severity = v$severity)
}
watch <- function(d) {
    d$median_detection <- round(d$median_detection, 3)
    d$zero_rate <- round(d$zero_rate, 3)
    d$share <- round(d$share, 3)
    d
}
tb <- do.call(rbind, lapply(c(0.002, 0.005, 0.01, 0.02, 0.05, 0.15, 0.4, 0.8),
                            gradient))
watch(tb)
```

Read it down the `share` column: at 0.2% detection most of the set's chance of
entering the score is tie-break, and at 15% almost none of it is -- an order of
magnitude of movement over rows where the only thing that changed is how often
the cells detected the set. No pathway, no co-expression, no biology moved.

Three criteria can raise a set's severity, and the table shows them crossing
independently of each other. The tie-break share drops below its moderate
threshold at 2% detection, while the zero rate is still 0.75 and stays in the
high band down to 5%: the share asks how much of the score's *chance* came from
the ranking, and the zero rate asks how many *cells* carry any information
about the set at all. The third is the gap between the observed zero rate and
the rate the scoring rule implies under independence, and it is why the bottom
rows are not the clean bill of health they look like. Those sets are detected
in enough cells to have almost no tie-break component, and they are flagged
anyway, because genes that are found together also depart from the independence
the conventional null assumes -- in the other direction. That is the case the
co-detection-matched null is for.  The `reasons` field of the verdict carries
one line per criterion that fired.

The verdict is also not a property of the gene set alone. Rerun this on a
dataset with deeper cells and the same rows shift right, because the tie-break
enters the score only where a cell's depth falls short of the ceiling.

The rest of the vignette uses a set from the middle of that table, where the
answer is genuinely contested.

```{r}
gene_set <- set_at_detection(0.02)
```

## Step 2: a null the data supports

A permutation test that shuffles cells answers "is this score different from
what the same cells would give by chance", which is not the question. The
question is whether the *set* carries information beyond its detection
profile, so the null has to be drawn from genes with the same profile.

`construct_expression_matched_null()` draws replacement sets from the
detection-rate stratum of each gene of the set, excluding the whole query set
from the pool. It reports the composition it achieved, because a null that
silently failed to match is worse than no null at all.

```{r}
built <- construct_expression_matched_null(cache, gene_set, n_sets = 20,
                                           seed = 42)
data.frame(
    quantity = c("target detection", "null detection", "ratio",
                 "mean per-set gap", "worst per-set gap",
                 "draws within tolerance", "genes shared with the query",
                 "draws that had to relax the pool", "status"),
    value = c(round(built$target_detection, 4),
              round(built$null_detection, 4),
              round(built$detection_ratio, 4),
              round(built$mean_detection_gap, 4),
              round(built$worst_detection_gap, 4),
              round(built$frac_within_tolerance, 3),
              built$max_overlap, built$draw_relaxations, built$status))
```

Two of those columns are the ones to look at. `max_overlap` has to be zero: a
null set that reuses a gene of the query is not a null, it is the observation
with extra steps. And `status` is `FAIL` when the pool could not supply a set
close enough in detection rate, which happens when the query is so sparse that
nothing in the background resembles it -- a real finding about the data, not a
hiccup in the function.

`status` reads the *mean* gap rather than the worst one, and the two are both
reported because they answer different questions. A maximum over draws can only
grow as more draws are asked for, so a rule based on it would pass a family at
25 draws and fail the same family at 100 -- the diagnostic would punish the
better practice. The mean settles down instead, and `draws within tolerance`
shows how much spread sits behind it.

Now score the observation and its null on the same cells and correlate both
against an axis. The axis here is the depth itself, which is the worst case and
therefore the honest one to demonstrate with.

```{r}
axis <- cache$depth
cal <- compute_calibrated_score(cache, gene_set,
                                null_genes = built$null_genes,
                                axis = axis)
str(cal$summary[c("observed", "null_mean", "null_sd", "p_value")])
```

The observed correlation of the score with depth is compared against the
distribution of the same correlation over the matched null draws. A score that
tracks depth because the tie-break does will sit inside that distribution and
the p-value will be unremarkable; a score that tracks depth through its genes
will not. Here the observed value sits
`r sprintf("%.1f", abs(cal$summary$observed - cal$summary$null_mean) / cal$summary$null_sd)`
null standard deviations from the null mean, which is the quantity to report.
The naive test on the same score, which permutes the axis instead of matching
the genes, gives a p-value of
`r sprintf("%.3f", spearman(cal$observed, axis)$p)` -- and that comparison is
the reason the framework exists.

## Step 3: what the rule you were going to use actually does

The other silent source of false positives is the cutpoint. Choosing the
cutpoint that best separates the outcome and then testing it as if it had been
chosen in advance is the standard move in this literature, and it is
anticonservative in a way that is easy to measure. `cutpoint_fpr()` permutes
the outcome, re-applies the rule, and counts how often it declares a
significance it was never entitled to.

```{r}
outcome <- axis > stats::median(axis)
rules <- c("median", "optimum")
fprs <- lapply(rules, function(m) {
    cutpoint_fpr(cal$observed, outcome, n_perm = 300, method = m, seed = 7)
})
names(fprs) <- rules
data.frame(rule = rules,
           fpr = round(vapply(fprs, function(f) f$fpr, 0), 3),
           observed_p = signif(vapply(fprs, function(f) f$observed_p, 0), 2),
           cutpoint = signif(vapply(fprs, function(f) f$observed_cutpoint, 0), 3))
```

The median rule holds its level: its false-positive rate is about
`r ALPHA`, as a rule that never looked at the outcome should. The optimal rule
does not, and the multiple is what the outcome-driven cutpoint costs on this
score. Both rules report the *same* data on the same cells; they differ only in
whether the cutpoint was chosen by looking at the outcome. `n_perm = 300` here
for speed; the default is 1000.

## Step 4: the whole checklist in one call

`diagnostic_checklist()` runs all four steps on one gene set and returns a
serialisable object. This is the function to keep in a pipeline.

```{r}
check <- diagnostic_checklist(cache, gene_set, axis = axis, n_perm = 300,
                              label = "example set")
check$overall_verdict
check$severity
cat(check$reasons, sep = "\n")
```

The verdict is deliberately blunt. `NOT_IDENTIFIABLE` means the score cannot
carry the cell-level claim being made about it, whatever the p-value was;
`INTERPRETABLE_WITH_MATCHED_NULL` means the association may be real but every
test of it has to be run against a matched null, and the object carries the one
that was constructed. `to_json()` writes it out.

```{r}
path <- tempfile(fileext = ".json")
to_json(check, path)
cat(substr(readLines(path), 1, 60), sep = "\n")
unlink(path)
```

## What changes and what does not

The framework does not tell you that your association is false. It tells you
what part of the evidence for it came from the measurement and what part came
from the ranking, which is a distinction the conventional analysis does not
draw. Three things follow, in the order they are usually needed:

1. **Report the share.** A tie-break share of 0.6 means that 60% of the set's
   chance of entering the score comes from genes the cell did not express.
   That sentence belongs in the results whether or not it changes the
   conclusion.
2. **Replace the null.** Every test of the association -- correlation,
   group difference, survival -- should be run against expression-matched
   draws, and the p-value reported from that distribution rather than from the
   cells.
3. **Match the regime across cohorts.** A replication in a dataset with a
   different depth distribution tests a different score. `comparability_report()`
   reports whether two panels place the same set on the same scale before you
   read a replication failure as a biological disagreement.

One thing this does *not* license is treating every sparse set as
uninformative. A set can be sparse and still carry signal: the split in
`score_decomposition()` separates the two contributions cell by cell, and if
the detected part is doing the work, the sparsity is a caveat and not a
refutation.

```{r, fig.width = 6.5, fig.height = 3.6}
split <- score_decomposition(cache, gene_set, X$matrix > 0,
                             max_rank = ceiling)
# The cells that score at all, in fifths of depth, each point the mean of its
# fifth.  Two things are being kept out of the picture: the three-quarters of
# cells whose score is exactly zero, which carry no rank information either
# way, and the cell-to-cell spread within a fifth, which is wide enough to hide
# the trend if every cell were drawn.
scoring <- split$total > 0
tie_share <- split$tie_break[scoring] / split$total[scoring]
depth <- cache$depth[scoring]
fifth <- as.integer(cut(depth, stats::quantile(depth, seq(0, 1, 0.2)),
                        labels = FALSE, include.lowest = TRUE))
by_fifth <- function(v) as.numeric(tapply(v, fifth, mean))
okabe <- c("#0072B2", "#D55E00")
op <- par(bty = "n", mar = c(4.1, 4.3, 2.8, 1.2), las = 1,
          col.axis = "#333333", col.lab = "#333333", fg = "#333333")
plot(by_fifth(depth), 1 - by_fifth(tie_share), type = "b", pch = 19, lwd = 2,
     col = okabe[1], ylim = c(0, 1.18),
     xlab = "Genes detected in the cell", ylab = "Share of the cell's score",
     main = "Where the score came from, cell by cell")
lines(by_fifth(depth), by_fifth(tie_share), type = "b", pch = 19, lwd = 2,
      col = okabe[2])
# Labelled at opposite ends: the curves cross, so labels placed at the same
# depth would sit on top of each other.  The offsets are in the direction the
# label has room in, so neither runs off the margin.
n <- length(by_fifth(depth))
text(by_fifth(depth)[1], by_fifth(tie_share)[1], "tie-break fill",
     col = okabe[2], adj = c(0, -0.75), font = 2)
text(by_fifth(depth)[n], 1 - by_fifth(tie_share)[n], "measured genes",
     col = okabe[1], adj = c(1, -0.75), font = 2)
par(op)
```

The orange curve is the part of the score that the tie-break supplied. In the
shallowest fifth of the cells that score at all it is three quarters of the
score, and in the deepest fifth it is nothing -- which is exactly what makes it
a confound, since it is a monotone function of depth and depth is rarely the
variable of interest. The blue curve is the part the cells measured. Averaged
over the set, the orange is the `tie_break_share` of the table above.

## When the framework refuses, and when it only warns

Three inputs break the framework rather than stretching it, and the difference
between the two responses is deliberate. Each is shown here through the same
entry point an analyst would use, on a small matrix built to trigger it.

```{r}
set.seed(6)
n_cells <- 120
panel <- paste0("g", seq_len(200))
Xe <- matrix(0, n_cells, length(panel), dimnames = list(NULL, panel))
# Three blocks, and the two examples below read one each.  The first five genes
# are expressed in every cell and outrank everything else, so they hold ranks
# 1-5 in every cell and a set made of them scores the same number everywhere.
# The next twenty are switched on only in the second half of the cells, so a
# set made of *those* separates the two halves completely.  The rest are
# detected at a middling rate in every cell, which is the background both sets
# are drawn against.
Xe[, 1:5] <- 8
Xe[61:120, 21:40] <- 5
Xe[, 41:200] <- matrix(stats::rbinom(n_cells * 160, 1, 0.4), n_cells, 160)
Xe <- Matrix::Matrix(Xe, sparse = TRUE)
cache_e <- rank_cache(Xe, panel, ceiling = 10)
axis_e <- stats::rnorm(n_cells)
```

**A gene set with no spread.** If the set's genes are detected in no cell, or in
every cell, the score is the same number for every cell and its correlation with
any axis is 0/0. The framework stops rather than returning a `NaN` that would
surface much later, in whichever statistic first touched it, with nothing left
to say which set produced it.

```{r, error = TRUE}
diagnostic_checklist(cache_e, panel[1:5], axis = axis_e, n_perm = 20,
                     cutpoint_method = "median")
```

**A null that cannot be drawn.** Every draw excludes the query set, so what can
stand in for a 400-gene set is the background *minus those 400 genes*, not the
background. When fewer than `MIN_BACKGROUND_GENES` candidates are left, the
detection floor is lowered by 10% and the widening is announced; if that still
leaves too few, the call stops. A replacement drawn from five genes is not a
null, and the P value it produced would be read as a test.

```{r, error = TRUE}
construct_expression_matched_null(cache_e, panel[-1], n_sets = 1)
```

**A score that separates its groups almost perfectly.** This one is a warning
and not a refusal, because a marker that marks exactly what it is supposed to
mark also lands here, and a framework that withheld the result would be hiding
it rather than qualifying it. The three explanations worth checking travel with
the warning.

```{r}
seen <- character(0)
check_e <- withCallingHandlers(
    diagnostic_checklist(cache_e, panel[21:40], axis = as.numeric(1:n_cells),
                         n_perm = 20, cutpoint_method = "median"),
    sparsegs_event = function(w) {
        seen <<- c(seen, conditionMessage(w))
        invokeRestart("muffleWarning")
    })
# The record the warning produced, split back into its fields:
strsplit(grep("separation", seen, value = TRUE), " ")[[1]]
# and the number that produced it, which the checklist reports whether or not
# it crosses the threshold:
check_e$step3_false_positive_rate$score_auc
```

The separation is checked in both directions. The brief states the case as
"AUC >= 0.99", but `auroc()` does not fold its argument about 0.5, and a score
that lands every high-axis cell *below* every low-axis cell separates them just
as perfectly as one that does the reverse -- so reading only the upper tail
would have missed half the ways this confound can arise. The record carries
`auc` and `separation` separately for that reason.

## The log, and reading it as data

Every diagnostic above is a *record* rather than a sentence: an event name and a
set of named fields, written `event field=value` with any value containing a
space quoted. The point of the format is that a log can be counted and not only
read -- "how many of these sets had to be drawn from a widened pool" is a
question with an answer, and the answer is the reason the edge cases exist.

```{r}
# The call above ran at the default level, `warning`.  Wrapped so the records
# can be counted:
quiet <- function() {
    got <- character(0)
    withCallingHandlers(
        diagnostic_checklist(cache_e, panel[21:40],
                             axis = as.numeric(1:n_cells), n_perm = 20,
                             cutpoint_method = "median"),
        sparsegs_event = function(w) {
            got <<- c(got, conditionMessage(w))
            invokeRestart("muffleWarning")
        })
    got
}
length(quiet())
```

`set_log_level()` moves the floor for the framework's own records and for
nothing else. At `error` the same record is dropped before it is ever
formatted, so a caller who has decided the warnings are noise pays nothing for
them -- and the caller's own warnings are untouched either way.

```{r}
set_log_level("error")
length(quiet())
set_log_level("warning")
```

Records arrive as conditions of class `sparsegs_event`, which is what lets the
two be separated: `withCallingHandlers` above catches the framework's records
and re-raises everything else, so a `tryCatch(warning = ...)` around the same
call would see its own warnings and not ours.

The same records come out of the Python implementation of this framework, byte
for byte, and `experiments/cross_language_check.py` in the accompanying
repository asserts it: a claim about log compatibility that rests on reading two
files by eye is not a claim anyone can check.

## Reproducibility and where the numbers come from

Every function that draws anything takes a `seed`, and the defaults are stated
in `DEFAULT_CONFIG`. The tie-break deserves a note of its own. It is not a draw
from the random number generator: it is a key computed from the cell index, the
gene index and the seed, so the cache is identical whatever `chunk` is used and
identical to the Python reference implementation of the same framework, to the
integer. That is what lets the two implementations be tested against each other
rather than against each other's statistics.

```{r}
a <- rank_cache(X$matrix, X$genes, ceiling = ceiling, chunk = 400)
b <- rank_cache(X$matrix, X$genes, ceiling = ceiling, chunk = 37)
identical(a$clipped, b$clipped)
```

The full derivation, the rates the framework was calibrated on and the
comparison with the conventional analyses are in the accompanying manuscript.
`citation("sparseGenSetCal")` and the package documentation give the entry
points; `?sparsity_report`, `?construct_expression_matched_null` and
`?diagnostic_checklist` are the three to read first.

```{r}
sessionInfo()
```
